---
title: "FIPS / ICPSR"
output:
  github_document: default
---

## Setup

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
  
library(tidyverse)

```

We have two tables, one with ICPSR codes and one with FIPS codes: 

### ICPSR Table

```{r}
## ICPSR Excel table from https://usa.ipums.org/usa/volii/ICPSR.shtml
icpsr <- read_csv("data/icpsrcnt.csv") %>% 
  janitor::clean_names() %>% 
  rename(icp_county = county, 
         icp_state = state, 
         icp_statefips = statefips, 
         icp_statecode = stateicp, 
         icp_county_cod = county_cod) %>% 
  select(icp_state, icp_county, everything())

## ICPSR Table
icpsr
```


### FIPS table  

This is from my own repo. This may not be up to date! Can get more current list from the Census

```{r}
fips <- read_csv("https://raw.githubusercontent.com/kjhealy/fips-codes/master/county_fips_master.csv")

## FIPS table
fips
```



## Cleanup

We need to merge the tables but we don't have fully shared ID codes between them. We can try using the combination of State name and County name. The difficulty is that the counties (also e.g. parishes [in LA] and cities [in VA] etc) have the name "County" etc appended in one table but not the other. We'll strip the last word of the `county_name` column in the `fips` table to make them roughly comparable. This won't be perfect, though. 

```{r}
end_words <- c("Borough", "County", "Area", "Municipality", 
               "Columbia", "Parish", "city", "City")

## A regular expression to match them
end_words_re <- paste0(end_words, collapse = "|")
```

Now we add a column to `fips` that's the "county" names but with end word deleted.

```{r}
fips <- fips %>% 
  mutate(county_clipped = str_remove(county_name, end_words_re), 
         county_clipped = str_squish(county_clipped)) %>% 
  select(fips, county_name, state_abbr, state, county_clipped, everything())

## Table
fips
```

## Joining the tables

### 1. Left-join `icpsr` table to `fips` table

If we put `icpsr` on the left side of the join, this will preserve all rows in `icpsr` but any rows in `fips` not matched in `icpsr` won't be merged.

```{r}
## Result: 3,265 x 19
icpsr_ljoin_fips <- icpsr %>% 
  left_join(fips, by = c("icp_state" = "state_name", 
                         "icp_county" = "county_clipped"), 
            keep = TRUE)

write_csv(icpsr_ljoin_fips, file = "data/icpsr_ljoin_fips.csv")


icpsr_ljoin_fips

```

What's missing in this table after the merge? If FIPS is missing, the county didn't exist in the original `fips` table.

```{r}
## 279 rows missing
icpsr_nofips <- icpsr_ljoin_fips %>% 
  filter(is.na(fips)) 

icpsr_nofips

write_csv(file = "data/icpsr_nofips.csv", icpsr_nofips)
```

### 2. Left join `fips` to `icpsr`

If we put `fips` on the left side of the join, this will preserve all rows in `fips`, but any rows in `icpsr` not matched in `fips` won't be merged.

```{r}
# Result: 3,146 × 19
fips_ljoin_icpsr <- fips %>% 
  left_join(icpsr, by = c("state_name" = "icp_state", 
                           "county_clipped" = "icp_county"), 
             keep = TRUE)

write_csv(fips_ljoin_icpsr, file = "data/fips_ljoin_icpsr.csv")

fips_ljoin_icpsr
```

What's missing after merging this way? If the `icp_county_cod` is missing, the county didn't exist in the original `icpsr` table. 

```{r}
### 160 rows missing
missing_icpcod <- fips_ljoin_icpsr %>% 
  filter(is.na(icp_county_cod)) 

missing_icpcod

write_csv(missing_icpcod, file = "data/missing_icp_cod.csv")

```


### 3. Full join 

This is the union of the tables, preserving all rows in both tables, even if one isn't present in the other table.

```{r}
## Result: 3,425 x 19
county_fulljoin <- icpsr %>% 
  full_join(fips, by = c("icp_state" = "state_name", 
                                "icp_county" = "county_clipped"), 
            keep = TRUE)

write_csv(county_fulljoin, file = "data/county_fulljoin.csv")

county_fulljoin

```

The missings from both the original tables are in this table.



  
